---
title: "Question 3"
output: html_document
---
 
```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
```
 
# Goal
 
####Explore the relationship between ldl levels and attributes such as tobacco, obesity, adiposity and alcohol.We want to check if there is a strong correlation between ldl levels and those four variables often associater to higher risk of disease.
 
 
```{r}
load("heart.Rdata")
library(Amelia)
library(corrplot)
library(plyr)
library(ggplot2)
```

#Exploring very roughly the variable ldl and the dataset

####Missing data
Check if any data is missing
```{r}
missmap(heart)
```

There are no missing data.

####Boxplot

```{r}
boxplot(heart$ldl)
```

####Histograms

```{r}
hist(heart$ldl)
```

It is quite far from being gaussian.

####Boxplot log(ldl)

```{r}
boxplot(log(heart$ldl))
```

####Histogram log(ldl)
```{r}
hist(log(heart$ldl))
```

Looks gaussian, maybe we will use for a further analysis

```{r,echo=FALSE}
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```



#####Adding log(tobacco) and log(alcohol) to the dataframe
#####does it makes sense in your opinion? or i should just keep tobacco and alcohol?

```{r}
heart$logtob=log(heart$tobacco)
heart$logalc=log(heart$alcohol)
```

#Scatterplots

```{r}
p1=ggplot(data=heart,aes(x=heart$logtob,y=heart$ldl)) +
  geom_point()+
  xlab('log(tobacco)')+
  ylab('ldl')
p2=ggplot(data=heart,aes(x=heart$logalc,y=heart$ldl)) +
  geom_point()+
  xlab('log(alcohol)')+
  ylab('ldl')
p3=ggplot(data=heart,aes(x=heart$adiposity,y=heart$ldl)) +
  geom_point()+
  xlab('adiposity')+
  ylab('ldl')
p4=ggplot(data=heart,aes(x=heart$obesity,y=heart$ldl)) +
  geom_point()+
  xlab('obesity')+
  ylab('ldl')

multiplot(p1,p2,p3,p4,cols=2)
```

We can see tobacco and alcohol do not seem to affect relevantly the ldl level.
Instead higher levels of adiposity and obesity seem to be related with higher ldl levels. 



#Correlation plot

We display a matrix showing correlation between ldl,adiposity,obesity,tobacco and alcohol.

```{r}
heartred=as.data.frame(heart[,2:4],title=TRUE)
heartred$obesity<-heart$obesity
heartred$alcohol<-heart$alcohol

M<-cor(heartred)
corrplot(M,method="circle")
```

The most important positive correlations for ldl are adiposity and obesity.
We can also note that adiposity has a strong correlation with obesity and tobacco.

####Data analysis plan

One of the simplest tecniques to build a prediction model for a numerical response based on numerical predictors is linear modelling.
Since adiposity and obesity are strongly correlated we would omit one of the two from the model.

```{r}
#g=lm(log(ldl)~tobacco+alcohol,heart) wich one is better??
g=lm(ldl~tobacco+alcohol+obesity,heart)
summary(g)
shapiro.test(g$residuals)
```

