---
title: "ida_hmw_1.2.3"
author: "Mehdi"
date: "19 octobre 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Choosing a subset of 4 or 5 quantitative variables and explore linear relationships.

```{r}
library(moments)
library(corrplot)
cars = read.delim("/home/mehdi/Bureau/ida_hmw_1.2/cars.txt", sep = "", dec = ".")
```

For this part, we decided to choose 5 qualitative variables : mpg, displacement, horsepower, weight and acceleration
We created a new data set with only these 5 qualitative variables : carsquali

```{r}
carsquali=as.data.frame(cars[,3:6], title=TRUE)
carsquali$mpg <- cars$mpg
```

###Covariance and Correlation matrix

```{r}
cov(carsquali)
r=cor(carsquali)
r
```
We just displayed below the R-matrix. We can visualize it with a plot showing the correlation between the variables in a more intuitive way :
```{r}
corrplot(r, method = "circle")
```

Which variables are correlated the most ?

```{r}
diag(r)=0
which(r==max(abs(r)), arr.ind=TRUE)
```
The more correlated variables are weight and displacement which is logical because physically, displacement is strongly related to the car's weight.
Indeed, the correlation between these two variable is : 0.9128930 which is very high. We have a very strong linear relationship which implies that variables are measuring almost the same
thing. However, in absolute value, the correlation between all these variables apart between acceleration and weight (0.4985387) and between acceleration and mpg (0.4747480) are high. We have a strong relationship implying that we can be confident that these two variables are
connected in some way.

###Partial correlations

Let's explore the relationship between our 5 variables through partial correlations :

```{r}
library(ppcor)
pcor(carsquali)
matrix.partial=pcor(carsquali)$estimate
```

Visualizing partial correlations

```{r}
library(corrgram)
library(corrplot)
corrgram(matrix.partial,
         lower.panel=panel.shade, upper.panel=panel.pie,
         diag.panel=panel.minmax, text.panel=panel.txt)

corrplot.mixed(matrix.partial,order="AOE" )
```

Displaying the plots, we can see that partial correlations are closer to zero than ordinary correlations. In this case, this is suggesting that the relationship
between the variables of interest could be explained by their common relationships to the
explanatory variables upon which we are conditioning.

###Coefficient of determination

Define a function r2multv for squared multiple correlation coeficients

```{r}
r2multv<-function(x){
  r2s=1-1/(diag(solve(cov(x)))*diag(cov(x)))
  r2s
}
```

use it on data set "carsquali"

```{r}
r2multv(carsquali)
```


## R Markdown
