---
title: "Homework 1.3"
author: "Lorenzo Foa' Mehdi El Idrissi"
date: "06 novembre 2018"
output: html_document
df_print: paged
---

## Introduction


```{r, warning=FALSE,message=FALSE}
library(ggplot2)
library(GGally)
library(FactoMineR)
library(factoextra)
library(MASS)
library(dplyr)
```




Performing the PCA on the correlation matrix is equivalent to standardizing each of the variables. In general, PCA with and without standardizing will give different results. Especially when the scales are different. So since i have very dfferent scalses in my dataset I will use R matrix.

```{r}
Wines_complete = read.delim("/home/mehdi/Bureau/wines.txt", sep = "", dec = ".",header=FALSE,col.names=c('fixedacidity','volatileacidity','citicacid','residualsugar','chlorides','freesulfurdioxide','totalsulfurdioxide','density','pH','sulphates','alcohol','quality','type'))

Wines=select(Wines_complete,fixedacidity,volatileacidity,citicacid,residualsugar,chlorides,freesulfurdioxide,totalsulfurdioxide,density,pH,sulphates,alcohol,type)

Wines=na.omit(Wines)
```


#PCA
The main purpose of principal component analysis is to identify hidden pattern in a data set, to reduce the dimensionnality of the data by removing the noise and redundancy in the data and also to identify correlated variables.

```{r}
Wines_PCA_r=PCA(Wines, scale.unit = TRUE, ncp = 4, ind.sup = NULL, 
                quanti.sup = NULL, quali.sup = 12, row.w = NULL, 
                col.w = NULL, graph = F)

```


screeplot

```{r}
plot(1:length(Wines_PCA_r$eig[,1]),Wines_PCA_r$eig[,1],type = "o",main = "Scree plot",xlab = "Component number",ylab = "Eigenvalues")
plot(1:length(Wines_PCA_r$eig[,2]),Wines_PCA_r$eig[,2],type = "o",main = "Scree plot",xlab = "Component number",ylab = "Percentage of variance explained")
```

We can note that the eigenvalues are 'large' until the fourth/fifth component, the sixth starts being 'small'. These plots show us that in both cases, the eigenvalues and the percentage of variance explained are high for the fourth first components (eigenvalue > 1.2 and percentage > 10). In that sens, we decide to keep the first four components.


Analytical description of the first 4 principal components

```{r}
Wines_PCA_r$var$coord
```

Variability explained by the first principal component

```{r}
Wines_PCA_r$eig[1,2]
```

cumulative variance explained by the first 4 variables

```{r}
Wines_PCA_r$eig[1:4,2:3]
```


A graphic representation of the variables' contribute to the first two PCs.
For a given dimension, any row/column with a contribution above the reference line could be considered as important in contributing to the dimension.

```{r}
fviz_contrib(Wines_PCA_r, choice = "var", axes = 1, top = 11)
fviz_contrib(Wines_PCA_r, choice = "var", axes = 2, top = 11)
```



observation on the first 2 components by type

```{r}
fviz_pca_ind(Wines_PCA_r,habillage = "type", addEllipses = TRUE)

```

Crearly red and white wine is not the same!
but all the white wines have common charcteristics and red wines as well
The first principal component clearly separates wines by their type.
Can see some outliers. Their indexes are:15,67,95.

#interpretation pc1 and pc2
The first principal component can be interpreted as the inverse of the presence of

```{r}
Wines1=select(Wines_complete,fixedacidity,volatileacidity,citicacid,residualsugar,chlorides,freesulfurdioxide,totalsulfurdioxide,density,pH,sulphates,alcohol,quality)

```

```{r}
Wines_PCA_r1=PCA(Wines1, scale.unit = TRUE, ncp = 4, ind.sup = NULL, 
                quanti.sup = NULL, quali.sup = 12 , row.w = NULL, 
                col.w = NULL, graph = FALSE)
```

observation on the first 2 components by quality

```{r}
fviz_pca_ind(Wines_PCA_r1,habillage="quality")
```

As we can see form the plot, if we try to group by the quality, there is no pattern.

Lumping quality

```{r}
Wines1$quality[Wines1$quality<6]=1
Wines1$quality[Wines1$quality==6]=2
Wines1$quality[Wines1$quality>6]=3
```


PCA 

```{r}
Wines_PCA_r1=PCA(Wines1, scale.unit = TRUE, ncp = 4, ind.sup = NULL, 
                quanti.sup = NULL, quali.sup = 12, row.w = NULL, 
                col.w = NULL, graph = F)

```


```{r}
fviz_pca_ind(Wines_PCA_r1,habillage="quality")

```

Trying to separate the data between (more, less or equal than 6) quality's categories, we can also notice that there's no pattern in that case.

```{r}
fviz_pca_var(Wines_PCA_r1, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE)
```

For the variables : volatileacidity, chlorides, sulphates, fixedacidity, citicacid, freesulfudioxide, totalsulfurdioxide and Ph, cos2 is high because the variable are positioned close to the circumference of the correlation circle. This indicates a good representation of the variables on the principal component.