---
title: "Permutation test"
output: html_document
---
# Permutation test
The goal of this analysis is to investigate correlation between the variables Bill and Tip expressed as a percentage on the bill.

We are going to perform the permutation test on the hypothesis 
H0: cov(Bill,PctTip)==0 
                       vs.
H1: cov(Bill,PctTip)~=0


```{r setup, include=TRUE}
 knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(gtools)
library(combinat)
library(Rfast)
```
```{r}
load("C:/Users/utente/Desktop/Intelligent Data Analysis/RestaurantTips.Rdata")
```
####Exploring the data's distribution

```{r}
Rest<-RestaurantTips
hist(Rest$PctTip,breaks = 30)
hist(Rest$Bill,breaks = 30)
```


Compute the Pearson correlation coefficient
```{r}
co=cor(Rest$Bill,Rest$PctTip,method='pearson')
co
```

Look at the scatterplot Bill vs. PctTip
```{r}
ggplot(data=Rest,aes(x=Rest$PctTip,y=Rest$Bill)) +
  geom_point()+
  xlab('PctTip')+
  ylab('Bill')
```

Can not see a lot.



###R Pre-implemented permutation test

```{r}
#I set a seed because of randomness of the test
#set.seed(7)
nreps=10000
permcor(Rest$Bill,Rest$PctTip,R=nreps)
```

###Permutation test

```{r}
r.obt=co
y_matrix <- matrix(Rest$PctTip, nrow=nreps, ncol=length(Rest$PctTip),byrow=TRUE)
# need to transpose the matrix after calling the 'apply' function
# wondering why it change the matrix shape
y_matrix_shuffled <- t(apply(y_matrix, MARGIN=1, FUN=function(x) permute(x))) 
corr_vector <- apply(y_matrix_shuffled, MARGIN=1, FUN = function(x) cor(x, Rest$Bill))

prob <- length(corr_vector[abs(corr_vector) >= abs(r.obt)])/nreps

cat("P-value of the permutation test ",prob)


```

The p-value is quite high, so i will accept H0 as true, so the two variables ar uncorrelated.

Histogram of the distribution of r among the simulations, the red line stays for the Pearson correlation coefficient
```{r}
hist(corr_vector, breaks = 50, main =  expression(paste("Distribution around ",rho, "= 0")), xlab = "r from randomized samples")
a=c(co,co)
b=c(co,1000)
lines(a,b,col='red')

```


##Now i repeat the analysis whitout the outliers (percentage tip>30%)
```{r}
Rest<-Rest[Rest$PctTip<30,]
```

####Exploring the data's distribution

```{r}
hist(Rest$PctTip,breaks = 30)
hist(Rest$Bill,breaks = 30)
```


Compute the Pearson correlation coefficient
```{r}
co=cor(Rest$Bill,Rest$PctTip,method='pearson')
co
```

Look at the scatterplot Bill vs. PctTip
```{r}
ggplot(data=Rest,aes(x=Rest$PctTip,y=Rest$Bill)) +
  geom_point()+
  xlab('PctTip')+
  ylab('Bill')
```

Can see a weak correlation.



###R Pre-implemented permutation test

```{r}
#I set a seed because of randomness of the test
#set.seed(7)
nreps=10000
permcor(Rest$Bill,Rest$PctTip,R=nreps)
```

###Permutation test

```{r}
r.obt=co
y_matrix <- matrix(Rest$PctTip, nrow=nreps, ncol=length(Rest$PctTip),byrow=TRUE)
# need to transpose the matrix after calling the 'apply' function
# wondering why it change the matrix shape
y_matrix_shuffled <- t(apply(y_matrix, MARGIN=1, FUN=function(x) permute(x))) 
corr_vector <- apply(y_matrix_shuffled, MARGIN=1, FUN = function(x) cor(x, Rest$Bill))

prob <- length(corr_vector[abs(corr_vector) >= abs(r.obt)])/nreps

cat("P-value of the permutation test ",prob)


```

Since the p-value is very low, we have statistical evidence that the two variables ar correlated.

Histogram of the distribution of r among the simulations, the red line stays for the Pearson correlation coefficient
```{r}
hist(corr_vector, breaks = 50, main =  expression(paste("Distribution around ",rho, "= 0")), xlab = "r from randomized samples")
a=c(co,co)
b=c(co,1000)
lines(a,b,col='red')

```

