---
title: "Homework 2.1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Diamonds dataset

```{r, warning=FALSE,message=FALSE}
library(knitr)
library(gridExtra)
library(ggplot2)
library(car)
library(tseries)
library(lmtest)
```

```{r, results='asis', echo=TRUE}
diamonds <- read.table("HW-diamonds.txt", header=FALSE, col.names = c('caratage', 'color_purity', 'clarity', 'cert', 'price'))
kable(diamonds[1:5,])
```

## Caratage (Q1)

```{r}
G1 <- ggplot(diamonds, aes(x=caratage, y=price)) + geom_point() + geom_smooth(method = "lm")
G2 <- ggplot(diamonds, aes(x=caratage, y=log(price))) + geom_point() + geom_smooth(method = "lm")
grid.arrange(G1,G2,ncol=2)
```

From these plots it seems that there is a more linear relationship between caratage and log(price), than between caratage and price. Therefore, we choose to use log(price) as our response variable.

## Categorical information (Q2)

```{r}
# reorder
diamonds$color_purity = relevel(diamonds$color_purity, ref="I")
diamonds$clarity      = relevel(diamonds$clarity, ref="VS2")
diamonds$cert         = relevel(diamonds$cert, ref="HRD")

# fit model
lm1 = lm(formula = log(price) ~ caratage + color_purity + clarity + cert, data=diamonds)
summary(lm1)
```
### Interpretation of model fit

* Besides certGIA we find high significant net effects on all explenatory variables.
* One point increase in caratage increases the `log(price)` 2.855 times, everything else being the same. A 2.855 increase in `log(price)` is equivalent to a `exp(2.855) = 17.374` increase in price.
* Having color purity D increases the `log(price)` 0.42 times, compared to having color purity I, everything else being the same. A 0.417 increase in `log(price)` is equivalent to a `exp(0.417) = 1.517` increase in price.
* Having clarityVVS1 influences the price almost just as much as having clarityIF, everything else being equal.
* Having certIGI is worse for price than certHRD. Having certIGI compared to certHRD decreases the price `1 - exp(-0.174) = 0.160` times.

### Model plots

```{r}
par(mfrow=c(2,2))
plot(lm1, which=c(1:4), ask=F)
```

From these plots we can already see that the variance is not constant. It first increases when the predicted values increase and later decreases again. Also, the residuals seem to not be normally distributed. They look rather uniformly distributed. In the Cook's distance plot we spot a few true outliers (110, 214, 223) and high magnitudes in general.

### Normality

```{r}
jarque.bera.test(lm1$residuals)
```
Testing statistically for residual normality we find indeed that the residuals are not normally distributed. Hence, the p-value is < 0.05 and so we reject the null hypothesis stating that the data is normally distributed.

```{r}
outlierTest(lm1, cutoff=0.05)
jarque.bera.test(lm1$residuals[-c(211)])
```
When we remove the one outlier detected by the outlier test, the p-value decreases even further.

### Constant variance

```{r}
bptest(lm1)
```

Testing for constant variance we find that we can reject the null hypothesis stating that the residuals are homocedastic. Hence, the test points out that we do not have constant variance in the residuals. 


### Independence of the residuals

```{r}
lag.plot(lm1$residuals)
dwtest(lm1)
```

Looking at the autocorrelation plot we assume that there is no independence in the residuals. This assumption is confirmed by the Durbin-Watson test by rejecting the null hypothesis stating that the autocorrelation of the residuals is equal to 0.


## Squared carat (Q 3b)

Let us see what happens if we include the squared caratage in the linear model. 

```{r}
lm3 = lm(formula = log(price) ~ caratage + I(caratage^2) + color_purity + clarity + cert, data = diamonds)
summary(lm3)
```

From the summary we see that the squared caratage is actually of significance for the model. Let us have a look at the model plots. 

### Model plots

```{r}
par(mfrow=c(2,2))
plot(lm3, which=c(1:4), ask=F)
```

From these plots it looks at least like the variance is constant. Also, the residuals seem more normal distributed, but it is unclear from the plot whether it really is. Let us look at the statistical test to quantify our assumptions.

### Statistical tests

```{r}
jarque.bera.test(lm3$residuals)
bptest(lm3)
dwtest(lm3)
```

From these tests we presume that the residuals are normally distributed, constant in variance, but not independent. 

## Comparing remedial actions (Q4)
Looking at both remdial actions we note that both improve the validity of the assumptions. Namely, in both cases the residuals are now normally distributed. However, adding the squared caratage also improves the constant variance assumption, while adding the clustered caratage groups does not. Hence, we have a better model fit in the squared caratage case.

On the other hand, adding a squared term to the model complicates its interpretability. For example, in this case, when the caratage raises by one point, the price increases `exp(5.671)=290` times, while increasing the squared caratage by one point increases the price `exp(-2.103)=0.122` times (i.e. decreases 88%). This makes it hard to assess the influence of the caratage on the price.

The conclusion is that the comparison of the remedial actions is a comparison of model fit vs. interpretability. How this trade-off should be made depends on the application of the regression model. Since we are trying to educate laypeople, we believe that interpretability should be valued higher in this case. Therefore, we prefer the first remedial action.